Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-06-04 main.tex
Improvements to measurements section.

\begin{figure}
  \centering
  \subcaptionbox{}{
    \includegraphics[width=70mm]{figures/uncontrolled-eig-vs-speeds.png}
  }
  \subcaptionbox{}{
    \includegraphics[width=70mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
  }
  \caption{Uncontrolled (a) and controlled (b) root locus of eigenvalue
    components (real: solid, imaginary: dashed) plotted versus speed. Vertical
    blue lines indicate the two speeds we perform experiments at. Green shaded
    region is the linear stable range.}
  \label{fig:eigenvalues}
\end{figure}

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
at the handlebar via four inline load cells, each with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
to that described in the thesis include simplifying the controller with a
inexpensive microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
\begin{figure}
  \centering
  % TODO : scale these pngs to a reasonable size for an A4 sheet
  \subcaptionbox{
    A participant on the bicycle in the safety harness with the Bump'Em motor
    lines attached to the ends of the handlebars.
    \label{fig:participant-in-set-up}
  }{
    \includegraphics[width=75mm]{figures/participant-in-set-up.png}
  }
  \subcaptionbox{
    Top view diagram of bicycle handlebar perturbation system.
    \label{fig:setup-diagram}
  }{
    \includegraphics[width=65mm]{figures/setup-diagram.png
  }
}
\end{figure}

@@

%
\begin{table}
  \caption{Raw measurements}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Perturbation Order Number & \(j\) & integer & NA\\
    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
    \bottomrule
  \end{tabular}
\end{table}

Based on the findings in \cite{Haitjema2023a}, we calculate several variables.
We use the angular impulse \(L\) due to the perturbation forces over a
0.3~\si{\second} duration to characterize the magnitude of delivered
perturbation. The duration is selected based on the duration of the commanded
force.

\begin{align}
  F_r = & F_{rf} - F_{rr} \\
  F_l = & F_{lf} - F_{lr} \\
  L = & \int_{0\si{\second}}^{0.3\si{\second}} l/2(F_r + F_l) dt
\end{align}

We characterize the skill of the rider by calculating the standard deviation of
the steer angle \(\sigma_\delta\) over the 90~\si{\second} period they were
asked to maintain their position at the center of the treadmill. At the
initiation of each perturbation we log the steer and roll angles to characterize
the configuration of the bicycle when perturbed. The gain setting on the
balance assist controller indicates if the assistance is off \(k=0\) or on at
two different levels: low \(k=8\) or high \(k=10\). A recovery from the
perturbation is successful if the person neither put their foot down onto the
treadmill surface or either wheel of the bicycle exits the width of the
treadmill belt. We record this as a binary variable \(f\).

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/data-processing.png}
  \caption{Thesis Fig 3.13 on pg 27}
\end{figure}

@@

\begin{table}
  \begin{tabular}{lll}
    \(L\) & \si{\newton\meter\second} & angular impulse of the applied
    perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(\sigma_\delta\) & standard deviation of steer angle during Phase X \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (6 or 10) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
  \end{tabular}
\end{table}

@@

\begin{figure}
  \includegraphics{example-image-a}
  \caption{Box plot or distrubtion plot showing graphical view of stats
  tables.}
\end{figure}

>>>>

\begin{figure}
  \centering
  \subcaptionbox{}{
    \includegraphics[width=100mm]{figures/uncontrolled-eig-vs-speeds.png}
  }
  \subcaptionbox{}{
    \includegraphics[width=100mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
  }
  \caption{Uncontrolled (a) and controlled (b) root locus of eigenvalue
    components (real: solid, imaginary: dashed) plotted versus speed. Vertical
    blue lines indicate the two speeds we perform experiments at. Green shaded
    region is the linear stable range.}
  \label{fig:eigenvalues}
\end{figure}

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
at the handlebar via four inline load cells, each with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
to that described in the thesis include simplifying the controller with a
inexpensive microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
\begin{figure}
  \centering
  % TODO : scale these pngs to a reasonable size for an A4 sheet
  \subcaptionbox{
    A participant on the bicycle in the safety harness with the Bump'Em motor
    lines attached to the ends of the handlebars.
    % TODO : This label causes a compile error. I think this is where it is
    % supposed to go.
    %\label{fig:participant-in-set-up}
  }{
    \includegraphics[width=75mm]{figures/participant-in-set-up.png}
  }
  \subcaptionbox{
    Top view diagram of bicycle handlebar perturbation system.
    %\label{fig:setup-diagram}
  }{
    \includegraphics[width=65mm]{figures/setup-diagram.png}
  }
  \label{fig:experimental-setup}
\end{figure}

@@

Based on findings from measuring riders without balance
assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
may influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:

\begin{align}
  F_r = & F_{rf} - F_{rr} \\
  F_l = & F_{lf} - F_{lr} \\
  L = & \int_{0\si{\second}}^{0.3\si{\second}} l/2(F_r + F_l) dt
  \textrm{.}
\end{align}

At the initiation of each perturbation we log the steer and roll angles to
characterize the configuration of the bicycle when perturbed. The gain setting
on the balance assist controller indicates if the assistance is off \(k=0\) or
on at two different levels: low \(k=8\) or high \(k=10\). A recovery from the
perturbation is successful if the person neither put their foot down onto the
treadmill surface or either wheel of the bicycle exits the width of the
treadmill belt. We record this as a binary variable \(f\). All measured
variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & \si{\meters\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\cite{Gabriel2023} \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

\begin{table}
  \small
  \begin{tabular}{lll}
    \toprule
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of the applied
    perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(\sigma_\delta\) & standard deviation of steer angle during Phase X \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (6 or 10) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics{example-image-a}
  \caption{Box plot or distrubtion plot showing graphical view of stats
  tables.}
\end{figure}
